%
% Copyright © 2017 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%
%----------------------------------------------------------------------------------------
%\part{Lecture notes}
   \mychapter{Introduction.}
      \input{L1introduction.tex}
%      \section{Problems}
   \mychapter{Mathematical background.}
      \input{convexOptimizationLecture2.tex}
      \section{Problems.}
         \input{ProblemSet1Problem1.tex} % Taylor series expansion
         \shipoutAnswer
         \input{ProblemSet1Problem2.tex} % Inversion formula for ``small'' matrices
         \shipoutAnswer
   \mychapter{Matrix inner product, SVD, and set types.}
      \input{convexOptimizationLecture3.tex}
      \section{Problems.}
         \input{ProblemSet1Problem3.tex} % Matrix inner product
         \shipoutAnswer
   \mychapter{Sets and convexity.}
      \input{convexOptimizationLecture4.tex}
      \section{Problems.}
         \input{ProblemSet1Problem4.tex} % Convex, affine, and conic hulls
         \shipoutAnswer
         \input{ProblemSet1Problem5.tex} % Distance between two parallel hyperplanes
         \shipoutAnswer
   \mychapter{Sets, epigraphs, quasi-convexity, and sublevel sets.}
      \input{convexOptimizationLecture5.tex}
      \section{Problems.}
         \input{ProblemSet1Problem6.tex} % Ellipses, eigenvalues, eigenvectors, and volume
         \shipoutAnswer
         \input{ProblemSet1Problem7.tex} % Proving convexity-preserving operations
         \shipoutAnswer
         \input{ProblemSet1Problem8.tex} % Expanded and restricted sets
         \shipoutAnswer
   \mychapter{First and second order conditions.}
      \input{convexOptimizationLecture6.tex}
      %\section{Problems}
   \mychapter{Examples of convex and concave functions, local and global minimums.}
      \input{convexOptimizationLecture7.tex}
      \input{convexOptimizationLecture8.tex}
      %\section{Problems}
   \mychapter{Compositions of functions.}
      \input{convexOptimizationLecture8Composition.tex}
   \mychapter{Problem set II (not attempted).}
      \input{ProblemSet2Problem1.tex}
      \input{ProblemSet2Problem2.tex}
      \input{ProblemSet2Problem3.tex}
      \input{ProblemSet2Problem4.tex}
      \input{ProblemSet2Problem5.tex}
      \input{ProblemSet2Problem6.tex}
      \input{ProblemSet2Problem7.tex}
      \input{ProblemSet2Problem8.tex}
      \input{ProblemSet2Problem9.tex}
      \input{ProblemSet2Problem10.tex}
