%
% Copyright © 2017 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%
\makeproblem{Maximum likelyhood estimation.}{convexOptimization:problemSet2:5}{
In this problem, we are given a set of data points \((x_i,y_i)\), \( i = 1 \cdots 100\). We wish to fit a quadratic
model,

\begin{equation}\label{eqn:ProblemSet2Problem5:20}
y_i = a x_i^2 + b x_i + c + n_i,
\end{equation}
to the data. Here, \((a, b, c)\) are the parameters to be determined and
\( n_i \) is the unknown observation noise. The \((x_i, y_i)\) points are contained in a file \matlabText{dataForMLest.mat}
available on the course webpage.  You may load the data to \matlabText{MATLAB} using the command
\matlabText{load ps01data}
and view them using \( \matlabText{scatter(x,y,'+')}\). Please use the same data set and find the
maximum likelihood estimate of \((a, b, c)\) assuming \(n_i\)'s are i.i.d. when

\makesubproblem{}{convexOptimization:problemSet2:5a}
\( n_i \sim N(0, 1)\);
\makesubproblem{}{convexOptimization:problemSet2:5b}
\(n_i\) is always positive and \( p_{n_i}(z) = e^{-z} u(z)\) where \(u(\cdot)\) is the unit step function.

Please plot the data and the models on the same \matlabText{MATLAB} figure and submit the figgure as a part of
your solution.
(\matlabText{MATLAB} has built-in functions to solve many optimization problems. For example, \matlabText{linprog}
solves a linear programming problem, quadprog solves a quadratic programming problem. You
may use \matlabText{help linprog} to get more details.

\paragraph{Hint:} \partref{convexOptimization:problemSet2:5a} has an analytic solution.)
} % makeproblem

\makeanswer{convexOptimization:problemSet2:5}{
\withproblemsetsParagraph{
\makeSubAnswer{}{convexOptimization:problemSet2:5a}

TODO.
\makeSubAnswer{}{convexOptimization:problemSet2:5b}

TODO.
} % redaction
} % answer
