%
% Copyright © 2017 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%
\makeproblem{First and second order conditions for convexity}{convexOptimization:problemSet2:6}{
In class we proved the first and second-order conditions for convexity of differentiable scalar functions. In particular, the first-order condition we showed is that a differentiable function \( f : \bbR \rightarrow \bbR \) is convex if and only if \( \dom f \) is a convex set and

\begin{equation}\label{eqn:ProblemSet2Problem6:20}
f(x) \ge f(x_0) + f'(x_0) (x - x_0),
\end{equation}

for all \( x, x_0 \in \dom f\).
The second-order condition we showed is that a twice differentiable function \( f : \bbR \rightarrow \bbR \)
is convex if and only if \( \dom f \) is a convex set and \( f''(x) \ge 0 \) for all \( x \in \dom f \).

In this problem you are asked to prove the two corresponding vector results:

\makesubproblem{}{convexOptimization:problemSet2:6a}
If \( f : \bbR^n \rightarrow \bbR \)
is differentiable then \( f \) is a convex function if and only if \( \dom f \) is a convex set
and

\begin{equation}\label{eqn:ProblemSet2Problem6:40}
f(\Bx) \ge f(\Bx_0) + \lr{ \spacegrad f(\Bx_0)}^\T (\Bx - \Bx_0),
\end{equation}

for all \( \Bx, \Bx_0 \in \dom f \).

\makesubproblem{}{convexOptimization:problemSet2:6b}
If \( f : \bbR^n \rightarrow \bbR \)
is twice differentiable then \( f \) is a convex function if and only if \( \dom f \) is a
convex set and

\begin{equation}\label{eqn:ProblemSet2Problem6:60}
\spacegrad^2 f(\Bx) \ge 0,
\end{equation}
for all \( \Bx \in \dom f \).

To prove the above two results, follow the approach followed in class. Namely, show that the
differentiable (twice differentiable) vector function
\( f : \bbR^n \rightarrow \bbR \)
is convex if and only if the first
order (second-order) scalar condition holds along all lines in the domain.
In other words, show that
\( f \) is convex if and only if \( \dom f \) is a convex set and the first-order (second-order) scalar condition
holds for \( f(\Bx_0 + t \Bv) \) for all \( \Bx_0, \Bv \in \bbR^n, t \in \bbR \) and \( \Bx_0 + t \Bv \in \dom f \).

To be clear, we already proved part \partref{convexOptimization:problemSet2:6a}
in class. For \partref{convexOptimization:problemSet2:6a}
you are simply asked to reproduce
that proof to ensure you fully understand the proof method.
Clearly explain the overall logic and
the logic of each step.
Then, in \partref{convexOptimization:problemSet2:6b} you are asked to take the same (lines-based) approach to
show the second-order condition.
} % makeproblem

\makeanswer{convexOptimization:problemSet2:6}{
\withproblemsetsParagraph{
\makeSubAnswer{}{convexOptimization:problemSet2:6a}

TODO.
\makeSubAnswer{}{convexOptimization:problemSet2:6b}

TODO.
} % redaction
} % answer
